Does differentiation beget differentiation?

### 2022-10-28 Manuscript_3_draft1.Rmd
Newest version of paper.

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized at `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research. However, we still know very little about how attitudes towards the EU's differentiated integration are shaped by exposure to such differentiation in the past, and how this effect relates to individual Euroscepticism. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more differentiation are likelier to want a future EU that allows countries to opt out of undesirable policy integration. This effect is particularly strong among Eurosceptics. However, whether differentiated integration is voluntarily chosen or discriminatorily imposed seems to be less important. Altogether, my results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU. This potentially complicates the EU’s move towards an ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

@@

# Introduction
Despite the EU being one of the most advanced political cooperations at the global stage, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty. The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation despite contestation of it in groups of countries [@leuffen_differentiated_2013; @schimmelfennig_ever_2020]. Differentiation has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called “constitutional differentiation”. Second, “instrumental differentiation” has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules. Such differentiation can either happen voluntarily or because the EU's institutions impose differentiation as a precondition for membership [@schimmelfennig_instrumental_2014]. Thus, ‘variable geography’ [@stubb_categorization_1996] is in many ways a better description of the EU's political development than the union's own goal of "ever closer Union".

While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite_2022], we know little about how actual exposure to it has impacted support for the goal of a uniformly integrated EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain: First, as mentioned, we know too little about how past exposure to differentiation shapes support for it. Second, our understanding of how this relationship is moderated by Eurosceptic sentiments. Lastly, we do not know whether the voluntarily chosen or involuntarily imposed nature of past differentiation impacts citizens' preferences for differentiation as a future model of EU integration.

By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing country-level differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on individual-level Euroscepticism. However, there does not seem to be a clear connection between how the differentiated integration citizens have been exposed to in the past have been implemented and support for a less uniformly integrated EU. The greater support for differentiation among Eurosceptics can be explained by how such differentiation is more likely to bring about the less integrated Europe that such individuals may want. However, more research is needed to understand the seeming lack of a diverging response to differentiation even where it has been implemented under very different circumstances.

In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009] While the EU’s legal and functional reach is much greater than that of other international organizations, many of the organizations constituting the global governance system have seen their legal and political powers expand in recent years [@zurn_theory_2018]. Even if more research is needed to understand whether my results travel to contexts other than the EU, my results can help illuminate how the allowance of differentiations from an emerging international legal order can shape attitudes towards deepening transnational cooperation.

My findings are also important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to mitigate the challenges caused by public contestation [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009].

# Public support for European integration - what we know and what we don't
A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @gabel_public_1998; @hobolt_public_2015; @hooghe_calculation_2005]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or the social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is cue-based, with citizens relying informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018]. Lastly, strong identification with the nation-state potentially makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated issue is why some people support a more differentiated EU, in which the EU institutionally permits the divergence of the EU's legal order. Such differentiated integration is already comprehensively investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration has focused on liberal economic values [@leuffen_public_2022], Euroscepticism [de_blok_blessing_2021], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite_2022].

Thus, with one notable exception, the effect of contextual factors are under-studied. Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013]. One potential source of support for differentiated integration may be the varying exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer. If there is a misalignment between the actual and desired levels of integration, citizens may respond by either expressing greater or lesser support for a more differentiated EU that can bring about a political cooperation more aligned with their preferences [@busemeyer_beyond_2021; @wlezien_public_1995]. Previous literature has primarily investigated the potential feedback loops between EU policy-making, such as directives and regulations, and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020]. This paper instead focuses on how alterations to the EU's legal structure, in the form of negotiated or imposed exemptions from the EU's treaties, shapes attitudes towards more or less uniform integration.

Greater exposure to differentiated integration is also likely to make citizens more familiar with EU differentiation, thus increasing their knowledge of how differentiated integration would work in the future. Such knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015]. I also find it theoretically plausible that this relationship will depend on whether citizens support their countries' EU memberships or not: Because constitutional differentiated integration allows for permanent opt-outs from specific policy areas, Eurosceptics who favour leaving the EU are more likely to see differentiated integration as a viable path towards the less differentiated Europe that they primarily prefer. This yields two complementary hypotheses:

*H1a: Exposure to differentiated integration increases support for future differentiated integration.*

*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*

Responses to differentiated integration may also vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state governments. An important reason is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration. As @schimmelfennig_eu_2014 shows, countries in Central and Eastern Europe were among those exposed to the greatest amount of discriminatory differentiated integration. Among other things, the new Eastern member states were for several years excluded from Schengen, and thus the possibility of easily moving for work purposes. The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary. As figure A6 shows, citizens of Central and Eastern Europe are among those in this sample exposed to the greatest amount of differentiated integration.

Eurosceptics in these countries may also have more adverse reactions to differentiated integration than those supporting EU membership: Because citizens in these countries may predominantly be familiar with externally imposed and discriminatory differentiation, Eurosceptics may be more likely to identify a non-uniformly integrated EU that allows countries to opt out of integration with the potential for domination. Their concern for national sovereignty may again lead them to develop more strongly negative attitudes towards differentiated integration than what can be found among supporters of EU membership.

Elites in Central and Eastern Europe, especially in Poland, have also framed differentiated integration as a path towards a divided EU, with a powerful core and a less powerful periphery [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states outside the currency area would lose their influence over the EU's economic policy-making. As elite cues are particularly important for shaping attitudes towards less salient EU questions [@steenbergen_whos_2007], such as differentiated integration [@telle_elite_2022], citizens would be likely to respond to these discourses by expressing less support for differentiated integration. Thus, two different causal mechanisms could produce the same outcome.

However, this history might be less relevant if current political elites do not use it to shape preferences towards differentiated. Instead, the increasingly Eurosceptic attitudes expressed by Central and Eastern European politicians [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds. Exposure to such cues might, in contrast to what was stated before, make citizens more likely to express an increased demand for differentiated integration. This yields two competing hypotheses:

*H2a: Exposure to differentiated integration will lead to lower levels of support for future differentiated integration in Central and Eastern Europe than in the rest of the EU.*    
*H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*    
*H3: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.*

If exposure to discriminatory and externally imposed differentiation leads to less support for differentiated integration, especially among Eurosceptics, we are theoretically likely to find the opposite where differentiation has been voluntarily chosen. Scandinavia is an interesting case study of this theoretical assumption: Here differentiation has been chosen by national governments and justified by explicit references to how these differentiations enhanced national autonomy [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016]. Previous literature has also found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. In other words, it is likely that the familiarity that citizens in Scandinavia have with voluntarily chosen differentiation makes could make them more likely to support it as a future model of EU integration.

This is also particularly likely to be true for Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise favour. It may also be the case that they, because of a generally critical attitude towards EU integration, are more likely to see differentiated integration as a solution that would better facilitate the less integrated Europe they otherwise favour. I hypothesize the following:

*H4a: Exposure to differentiated integration will lead to greater levels of support for it among for citizens of Sweden and Denmark than in the EU as a whole.*

*H4b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*

My paper thus investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, as well as in Scandinavia and Central and Eastern Europe. I then test whether this relationship is moderated by pre-existing Eurosceptic sentiments. This lets me test both the general effect of exposure to differentiation as well as its interaction with pre-existing attitudes to European integration.

# Data and methods
```{r load data}
EUI_final_DI_disc <- readRDS("data/EUI_final_DI_disc.Rds")
nordic_data <- readRDS("data/nordic_data.Rds")
```

I investigate support for differentiated integration using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021]. The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`. The surveys include two questions about differentiated integration. The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration. This is called constitutional differentiation. The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration. This is an instrumentally differentiated EU. This article focuses on popular support for constitutional differentiation, as its allowance for an EU that is permanently segmented into multiple policy communities represents arguably the sharpest departure from the EU’s normative goal of an “ever closer Union”.

The survey data are combined with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2019. The data shows each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced. The data contain information about differentiations from EU treaties, or primary law, rather than differentiations from the less salient secondary law that constitutes the EU’s main legislative output [@duttle_opting_2017]. I believe the greater salience of primary law differentiation in public debates will lead to a greater knowledge of them among EU citizens than what is the case for differentiations from secondary law. They are thus more likely to be important for respondents as they make up their minds about the desirability of differentiated integration.

## Dependent variable
My dependent variable measures support for constitutional differentiated integration. The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU (“Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration. This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to cooperate. For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.)". The response categories are recoded, so that 1 becomes Strongly disagree and 5 becomes Strongly agree. By not asking about “constitutional differentiated integration”, the question becomes easier to understand. It also frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out.

## Variables of key interest
To operationalize each country’s exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force. This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced. I lastly log-transform the variable.

This operationalization has three benefits: First, it captures both the temporal and functional scope of a country’s exposure to differentiation. Second, the weighting scheme means that those instances of differentiation that were recently in force are given the most weight when investigating the impact of exposure to differentiated integration. This is also likely to correspond to how respondents relate their own experience of differentiation to support for it: While many respondents from Central and Eastern Europe may still remember the particulars of their countries’ exposure to externally imposed differentiation, possibly producing a cohort effect [@lauterbach_europe_2020], Southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and use this as a backdrop against which they shape their attitudes towards it. Finally, the log-transformation reduces the influence of extreme values, caused by particularly large numbers of differentiations in certain countries.

To analyse the regionally specific effects of differentiation, I create dummies for Scandinavian and Central and Eastern European citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. To create a Central and Eastern European dummy I do the same for everyone from Hungary, Lithuania, Poland, and Romania. These are used to investigate the interaction effects assumed by *H2-H4*. I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.

I operationalize support for liberal economic policies, which has been found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that “stresses economic integration, market competition and fiscal discipline”. This is contrasted to “a global Europe that acts as a leader on climate, human rights and global peace” or “a protective Europe that defends the European way of life and welfare against internal and external threats”. I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable. While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, I believe it still measures a general affinity for economic liberalism.

I include an equally weighted index of questions measuring respondents’ evaluation of the national economy. A factor analysis finds that these questions, in contrast to one asking about the respondent’s perceptions of the local job market, measure the same underlying dimension (see table A1). Previous literature shows that positive sociotropic economic evaluations correlate with support for both uniform and differentiated integration (see for instance @gabel_public_1998; @harteveld_europe_2013; @hooghe_calculation_2005; @leuffen_public_2022). Egotropic evaluations have, however, been shown to have little relevance for explaining support for differentiated integration [@leuffen_public_2022]. I measure support for national governments, relevant for explaining support for differentiated integration [@armingeon_loss_2014; @de_blok_blessing_2021; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation’s government is doing a good job.

I operationalize identity through a widely used question that asks respondents to rank the inclusiveness of their identities. I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0. While nested identities combining a national and European component are common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different nested identities [@hooghe_calculation_2005].

I also use ideology and gender as socio-political control variables. These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008). I operationalise ideology through a 7-unit scale, with 0 representing far-left and 6 far-right positions. As previous literature also finds evidence of a curvilinear correlation between ideology and support for the EU [@toshkov_beyond_2022; @van_elsas_changing_2015], I add a squared indicator of left-right ideology.

>>>>

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized on `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research. However, we still know do not know whether the increased familiarity with differentiated integration that exposure to it brings leads to greater support a differentiated EU, and whether the mode of implementation and attitudes towards the EU moderates this effect. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it. This particularly applies to Eurosceptics. It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy. My results suggest that exposure to the differentiated integration used to integrate highly salient policy areas can solidify popular support for a less integrated EU. This can complicate the EU’s move towards ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: TRUE
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

@@

# Introduction

Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty. The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called "constitutional differentiation". Second, "instrumental differentiation" has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules. Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014]. Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's political development than the EU's own goal of "ever closer Union".

While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023], we know little about how actual exposure to it has shaped support for the goal of a uniformly integrated EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain. This article tests three theoretically informed hypotheses: First, it tests whether support for differentiation leads to greater support for it on average. Second, it tests whether those more critical of EU membership become more positively inclined to differentiated integration if exposed to more of it. Lastly, it tests whether those citizens exposed to differentiated integration that was voluntarily chosen by their governments are more favourably inclined towards future differentiation.

By combining data from surveys fielded in 2020-2021 that measure citizens' attitudes towards constitutional differentiated integration in 13 EU member states with data detailing every country's exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism. I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty. This shows that familiarity with differentiated integration can lead to greater support for it in the future. However, more research is needed to understand what causes regional variations in support for differentiated integration.

In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009]. While the EU's legal and functional reach goes beyond that of most other international organization, many international organizations have seen their legal and political powers expand in recent years [@zurn_theory_2018]. Even if more research is needed to understand whether my results travel to contexts beyond the EU, they help illuminate how allowing countries to opt out of international rule-making shapes attitudes towards the future organization of the international legal order.

My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to reduce public contestation of integration [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU's institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009]. In this sense, the EU's differentiated integration may be both a help and hindrance for future integration: On the one hand, it may help facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013]. On the other, greater popular support for differentiated modes of integration can make it more difficult for the EU's institutions to get the public's buy-in to functionally necessary integration. The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.

# Public support for European integration - what we know and what we don't

A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @hobolt_public_2015]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or the social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is based on cues and framings, with citizens relying on informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @de_vries_euroscepticism_2018]. Lastly, strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated question is why some people support a more differentiated EU, which renders the EU's integration either permanently or temporarily non-uniform. Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration have focused on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022]. Thus, with one notable exception, the effect of national contexts are under-studied. Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013]. Here citizens' exposure to past differentiated integration plays a crucial role: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would preferences. The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995]. While there have been many studies of the feedback loop between public opinion and the EU's ordinary policy-making [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are in essence alterations to the EU's institutional structure.

It is reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015]. Similarly, familiarity with more differentiated ways of structuring EU memberships may lead to a stronger perception among citizens that EU memberships that allow countries to opt of policy integration may be less problematic.

This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership. Since constitutional differentiated integration allows for permanent opt-outs from policy integration, Eurosceptics who favour leaving the EU may see differentiated integration as a potentially more realistic path towards the less integration Europe that they prefer. This yields two hypotheses:  
*H1a: Exposure to differentiated integration increases support for future differentiated integration.*  
*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*

Responses to differentiated integration may also vary depending on the situation in which differentiated integration was imposed. It is for instance likely that citizens may become more favourably inclined towards differentiated integration if the form of differentiated integration they have been exposed to was a result of a conscious decision to opt out on the part of their own governments. The reason is that these decisions may be more likely to be seen by citizens as an exercise of their own national sovereignty within the remits of an EU membership.

Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016], and has typically been decided as a result of a referendum. Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. This shows that Scandinavian citizens may theoretically respond to differentiated integration, because of the particularly voluntary nature of how differentiated integration has been implemented in these regions, by wanting more differentiated integration in the future.

Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour. This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty. I thus hypothesize:

*H2a: Exposure to differentiated integration will lead to greater support for it among citizens of Sweden and Denmark than in the EU as a whole.*\
*H2b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*

My paper investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, and whether this relationship is moderated by pre-existing Euroscepticism. I then test whether the relationship differs according to whether citizens have been exposed primarily to voluntarily chosen or discriminatory differentiated integration.

# Data and methods

```{r load data}
EUI_final_DI_disc <- readRDS("data/EUI_final_DI_disc.Rds")
nordic_data <- readRDS("data/nordic_data.Rds")
```

I investigate support for differentiation using OLS models with country fixed effects and data from Yougov surveys fielded in 2020-2021 [@hemerijck_eui-yougov_2021]. The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`. The sample is drawn from the Yougov online panel, with quotas used to make the sample broadly representative of the populations from which they are drawn [@noauthor_yougov_nodate] The surveys include two questions about differentiated integration. The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration, so-called constitutional differentiation. The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration. This is an instrumentally differentiated EU. This article focuses on popular support for constitutional differentiation, whose allowance for what @schuessler_mapping_2023 define as sovereignty-enhancing differentiated integration represents the sharpest break with the EU's current goal of an 'ever closer Union'.

The survey data are combined with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2019. The data show each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced. I focus on differentiations from the EU treaties, so-called primary law, rather than exemptions from the less salient secondary law that constitutes the EU's main legislative output [@duttle_opting_2017]. The reason is that the greater salience of primary law differentiation makes it more likely to be well-known by citizens than more technical secondary legislation. Citizens are thus more likely to use it to make up their minds about the desirability of differentiation.

## Dependent variable

My dependent variable measures support for constitutional differentiated integration. The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU ('Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration. This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to cooperate. For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.'). I code the response categories so that 1 is Strongly disagree and 5 is Strongly agree. This question frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out. It thus primarily measures what @schuessler_mapping_2023 term sovereignty-enhancing differentiation.

## Variables of key interest

To operationalize each country's exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force. This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced. I lastly log-transform the variable.

This operationalization has three benefits: First, it captures both the temporal and functional scope of a country's exposure to differentiation. Second, the weighting scheme ensures that treaty exemptions that were recently in force are given the most weight when correlating exposure to differentiation to support for it. This is also likely to correspond to how the survey respondents use their own experiences to form attitudes towards differentiation: While respondents from central and eastern Europe may still remember the particulars of their countries' exposure to externally imposed differentiation, potentially leading to a cohort effect [@lauterbach_europe_2020], southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and relate this to their support for greater use of differentiation today. Finally, the log-transformation reduces the influence of outliers in the form of many differentiations from some countries.

To analyse the regionally specific effects of differentiation, I create dummies for Scandinavian and Central and Eastern European citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. These are used to investigate the interaction effects assumed by *H2*. I lastly create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0. While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).

## Control variables

I operationalize support for liberal economic policies, found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that "stresses economic integration, market competition and fiscal discipline". This is contrasted to "a global Europe that acts as a leader on climate, human rights and global peace" or "a protective Europe that defends the European way of life and welfare against internal and external threats". I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable. While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, it appears to still measure a general affinity for liberal economic policies.

I include an equally weighted index of questions measuring respondents' evaluation of the national economy and their local areas' employment opportunities. A factor analysis finds that these questions, in contrast to one asking about the respondents' perceived income relative to others their age and financial security, all measure sociotropic economic evaluations (see table A1). Previous literature shows that such evaluations correlate with support for differentiated integration. The same does not apply to egotropic economic evaluations [@leuffen_public_2022]. I measure support for national governments, relevant for explaining support for integration [@armingeon_loss_2014; @de_blok_blessing_2023; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation's government is doing a good job. As an indicator of income, I add an variable surveying respondents' perception of their own economic situation relative to others in their age group.

I operationalize national identities through a widely used question that asks respondents to rank the inclusiveness of their identities. I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0. While mixed national/European identities are most common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially territorially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different forms of mixed identities [@hooghe_calculation_2005].

I also use ideology, age and gender as socio-political control variables. These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008). I operationalize ideology through a scale where 0 represents far-left and 7 far-right positions. As previous literature also finds evidence of a curvilinear correlation between ideology and support for the EU [@toshkov_beyond_2022; @van_elsas_changing_2015], I add a squared indicator of left-right ideology.

```{r create descriptive data}
data_descriptives <- readRDS("data/Data_descriptives.Rds")
```
